Modelling the emergence of whisker barrels

### 2020-06-02 paper/briefpaper/brief.tex
Parameter sweep caption and methods updates
Methods needed updating with the newer competition terms.

\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig1.png}
    \caption{\textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
      by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
      measure departure (Honda-$\delta$; see \citealp{senft_mouse_1991}) from a
      perfect Voronoi tesselation. Right shows the initial distribution of axon
      branching density ($a$) for one thalamocortical projection, and two
      molecular guidance fields ($\rho$), \cmnt{where the domain $S$ has been
        traced from \textbf{A}}. \textbf{B} The strengths of interaction
      $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41
      projections by the lengths of green and blue arrows respectively, assuming
      that similar fields aligned to the posterior-anterior and medial-lateral
      axes in the ventroposterior medial nucleus of the thalamus are sampled at
      the locations of putative barreloid centers (reconstructed from
      \citealp{haidarliu_size_2001}). \textbf{C} Simulation results for parameters
      \cmnt{$N=41$, $\alpha=3.6$, $\beta=16.67$, $k=3$, $D=0.5$, $\gamma\in\pm 2$,
      $\epsilon=1.2$} and $\delta{t}=0.0001$. Colours indicate the thalamic
      projection for which the connection density is maximal and black lines
      delineate boundaries (see Movie
      S1). \textbf{D} Red dots show the Honda-$\delta$ metric obtained from
      simulation approaching that obtained from \cmnt{real} barrels in \textbf{A} (dotted
      line); black squares \cmnt{give a} measure \cmnt{of} the correspondence
      between the real and
      simulated barrel shapes, $\eta$ (the sum of differences in
      area; units \cmnt{mm$^2$});
      \cmnt{grey hexagons show the measure of localization, $\omega =
        \oiint_{S} \left( \mathrm{max}_i(c_i)/\sum_{i=1}^{N} c_i\right) \mathrm{d}S$. }
      \textbf{E} \cmnt{The localization measure plotted across the domain
        shows that the structure in the connection density which has developed
        by $t=30000$ is largely absent at $t=1000$. The
        colour map
        indicates the values of  $\mathrm{max}_i(c_i)/\sum_{i=1}^{N} c_i$. Overlaid
        contours show $c_i > 0.95 \mathrm{max}_i(c_i)$.} All scale bars 1mm.}
    \label{fig:main}
  \end{fullwidth}
\end{figure}

@@

The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching we
redefined
%
\begin{equation} \label{eq:comp}
  \color{blue}
  \chi_i(\mb{x}, t) = \frac{\epsilon a_i}{N-1} \nabla \textstyle\sum_{j\ne i}{a_j}
  \color{black}
\end{equation}
%
\cmnt{which contributes to the \emph{flux of axonal branching} as an
  additional diffusion-like term which makes the branching density for a given
  projection `flow away' from regions} where the branches of other projections
become dense. \cmnt{$\epsilon$ parameterizes the strength of this competitive
  interaction.} Note that this operation is local to individual afferent
projections.

@@

% Full page width
\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig2.png}
    \caption{\cmnt{\textbf{A} Parameter sweep caption goes here}}
    \label{fig:paramsweep}
  \end{fullwidth}
\end{figure}

@@

% This figure is text width
\begin{center}
  \includegraphics[width=\linewidth]{./Fig3.png}
  \captionof{figure}{Guidance
    fields and emergent barrel pattern in a Fgf8 misexpression experiment
    (c.f. \citealp{assimacopoulos_fibroblast_2012}), simulated by reflecting
    $\rho_1$ from Fig.\,\ref{fig:main}A at the join ($\epsilon=80$).}
  \label{fig:fgf8}
\end{center}

@@

The most involved part of this computation is to find the divergence of the
flux of axonal branching, $\mb{J}_i(\mb{x},t)$, the term in parentheses in
Eq.\,\ref{eq:da}:
%
\begin{equation}
  \label{eq:divJ}
  \nabla\vcdot\mb{J}_i(\mb{x},t) = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right).
\end{equation}
%
Note that the sum of the guidance gradients is time-independent and define
$\mb{g}_i(\mb{x}) \equiv \sum_{j=1}^{M} \gamma_{i,j} \nabla\rho_j(\mb{x})$.
Because the divergence operator is distributive, Eq.\,\ref{eq:divJ} can be
expanded using vector calculus identities (dropping references to $\mb{x}$ and
$t$ for clarity):
%
\begin{equation}
\nabla\vcdot\mb{J}_i = \nabla\vcdot\big(D \nabla a_i\big) - \nabla\vcdot\big(a_i \mb{g}_i\big).
\end{equation}
%
Applying the vector calculus product rule identity yields
%
\begin{equation} \label{eq:divJExpanded}
%
\nabla\vcdot\mb{J}_i =
%
D \dvrg a_i % term1
-
a_i\nabla\vcdot\mb{g}_i % term2
-
\mb{g}_i\vcdot\nabla a_i % term3
,
\end{equation}
%
which has three elements to compute: i)~$D \dvrg a_i$ (the Laplacian of
$a_i$); ii)~a time-independent modulator of $a_i$ (because
$\nabla\vcdot\mb{g}_i$ is a time-independent static field); and iii)~the
scalar product of the static vector field $\mb{g}_i$ and the gradient of
$a_i$. Each of the divergences can be simplified by means of Gauss's Theorem
following \cite{lee_hexagonal_2014}.

(i) The computation of the mean value of the Laplacian across one hexagon of
area $\Omega = \frac{\sqrt{3}}{2}d^2$, located at position $\mb{p}_0$, with
neighbours at positions $\mb{p}_1$--$\mb{p}_6$ is
%
\begin{equation} \label{eq:lapl}
\begin{split}
\langle D \dvrg a_i(\mb{p}_0,t) \rangle  = \frac{1}{\Omega} \oiint_\Omega \dvrg a_i(\mb{x}, t) \dif\Omega & = \frac{1}{\Omega} \oint \frac{\partial a_i}{\partial \hat{\mb{n}}} \dif\gamma \\
& \approx \frac{1}{\Omega} \sum_{j=1}^6 \frac{\partial a_i(\mb{p}_j)}{\partial \hat{\mb{n}}} \bigg\rvert_{\mathrm{mid}} v \\
& = \frac{2}{\sqrt{3} d^2} \sum_{j=1}^6 \frac{a_i(\mb{p}_j) - a_i(\mb{p}_0)}{d} \frac{d}{\sqrt{3}} \\
& = \frac{2}{3 d^2} \sum_{j=1}^6 \big(a_i(\mb{p}_j) - a_i(\mb{p}_0)\big),
\end{split}
\end{equation}
%
where $v = d/\sqrt{3}$ is the length of each edge of the hexagon and $\dif\gamma$
is an infinitesimally small distance along its perimeter.

ii) The computation of the second term in Eq.\,\ref{eq:divJExpanded},
$\langle a_i(\mb{p}_0,t)\nabla\vcdot\mb{g}_i(\mb{p}_0)\rangle$, can be written out similarly:

@@

iii) The final term in Eq.\,\ref{eq:divJExpanded} is the scalar product of two
vector fields which is straightforward to compute from their Cartesian
components.

By separating the computation of Eq.\,\ref{eq:divJ} into parts (i), (ii) \&
(iii), the no-flux boundary condition,
$\mb{J}_i(\mb{x},t)\big\rvert_{\mathrm{boundary}} = 0$, can be fulfilled. On
the boundary, the contribution to $\mb{J}$ resulting from the first term of
Eq.\,\ref{eq:divJExpanded} can be fixed to 0 by the `ghost cell method' in
which, during the evaluation of (i), a hex outside the boundary containing the
same value as the hex inside the boundary is imagined to exist such that the
flux of $\mb{J}$ across the boundary is 0. Then, $\mb{g}_i(\mb{x})$ can be
tailored so that it, and its normal derivative approach 0 at the boundary,
ensuring that the second and third terms of Eq.\,\ref{eq:divJExpanded} also
contribute nothing to $\mb{J}$. This is achieved by applying to
$\mb{g}_i(\mb{x})$ a sharp logistic function of the distance from $\mb{x}$ to
the boundary.

All code required to reproduce these results is available at
\url{https://github.com/ABRG-Models/BarrelEmerge/tree/eLife}. The
computations described in (i), (ii) and (iii) may be found in the class method
\code{RD\_James\cmnt{\_comp2}::compute\_divJ()} which calculates \code{term1}, \code{term2}
and \code{term3}, respectively.

@@

Movie corresponding to Fig.\,\ref{fig:main}C in the main paper. Simulation parameters were
$N=41$, $\alpha=3.6$, $\beta=16.67$, $k=3$, $D=0.5$, $\gamma\in\pm 2$,
$\epsilon=1.2$ and $\delta{t}=0.0001$. Colours indicate the thalamic
projection for which the connection density is maximal, black lines delineate
boundaries, and overlaid contours show $c>0.5$. The final frame in the movie
is step 25,000 of the simulation.

>>>>

\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig1.png}
    \caption{\textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
      by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
      measure departure (Honda-$\delta$; see \citealp{senft_mouse_1991}) from a
      perfect Voronoi tesselation. Right shows the initial distribution of axon
      branching density ($a$) for one thalamocortical projection, and two
      molecular guidance fields ($\rho$), \cmnt{where the domain $S$ has been
        traced from \textbf{A}}. \textbf{B} The strengths of interaction
      $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41
      projections by the lengths of green and blue arrows respectively, assuming
      that similar fields aligned to the posterior-anterior and medial-lateral
      axes in the ventroposterior medial nucleus of the thalamus are sampled at
      the locations of putative barreloid centers (reconstructed from
      \citealp{haidarliu_size_2001}). \textbf{C} Simulation results for parameters
      \cmnt{$N=41$, $\alpha=3.6$, $\beta=16.67$, $k=3$, $D=0.5$, $\gamma\in\pm 2$,
      $\epsilon=1.2$} and $\delta{t}=0.0001$. Colours indicate the thalamic
      projection for which the connection density is maximal and black lines
      delineate boundaries (see Movie
      S1). \textbf{D} Red dots show the Honda-$\delta$ metric obtained from
      simulation approaching that obtained from \cmnt{real} barrels in \textbf{A} (dotted
      line); black squares \cmnt{give a} measure \cmnt{of} the correspondence
      between the real and
      simulated barrel shapes, $\eta$ (the sum of differences in
      area; units \cmnt{mm$^2$});
      \cmnt{grey hexagons show the measure of localization, $\omega =
        \oiint_{S} \left( \mathrm{max}_i(c_i)/\sum_{i=1}^{N} c_i\right) \mathrm{d}S$. }
      \textbf{E} \cmnt{The localization measure plotted across the domain
        shows that the structure in the connection density which has developed
        by $t=30000$ is largely absent at $t=1000$. The
        colour map
        indicates the values of  $\mathrm{max}_i(c_i)/\sum_{i=1}^{N} c_i$. Overlaid
        contours show $c_i > 0.95 \mathrm{max}_i(c_i)$.} All scale bars 1\,mm.}
    \label{fig:main}
  \end{fullwidth}
\end{figure}

@@

The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching we
redefined
%
\begin{equation} \label{eq:comp}
  \color{blue}
  \chi_i(\mb{x}, t) = \frac{\epsilon a_i}{N-1} \nabla \textstyle\sum_{j\ne i}^{N}{a_j}
  \color{black}
\end{equation}
%
\cmnt{which contributes to the \emph{flux of axonal branching} as an
  additional diffusion-like term which makes the branching density for a given
  projection `flow away' from regions} where the branches of other projections
become dense. \cmnt{$\epsilon$ parameterizes the strength of this competitive
  interaction.} Note that this operation is local to individual afferent
projections.

@@

% Full page width
\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig2.png}
    \caption{\cmnt{\textbf{A} Exploring the parameter space of the model given
        in Eqs.\,1--3. Graphs show the three pattern metrics for $t=30000$
        steps as colourmaps: Honda $\delta$ (top row) shows conformance to a
        Voronoi pattern; $\eta$ (middle row) gives the simulated/experimental
        barrel area difference and $\omega$ (bottom row) is a measure of
        localization (separation) of the connection densities for the 41
        thalamocortical axon types. For each row, lighter (orange/yellow)
        colours imply a better model. The parameter space explored is three
        dimensional with the parameter ratio $\alpha/\beta$ varied along the
        columns of graphs, the competition parameter $\epsilon$ varied on the
        $y$-axes and the diffusion constant, $D$ varied along the $x$-axes. An
        asterisk (*) marks the parameter set displayed in
        Fig.\,\ref{fig:main}. Boxes i), ii) and iii) mark parameter sets for which
        corresponding patterns are shown in \textbf{B} (for $t=30000$ steps).}}
    \label{fig:paramsweep}
  \end{fullwidth}
\end{figure}

@@

% This figure is text width
\begin{center}
  \includegraphics[width=\linewidth]{./Fig3.png}
  \captionof{figure}{Guidance
    fields and emergent barrel pattern in a Fgf8 misexpression experiment
    (c.f. \citealp{assimacopoulos_fibroblast_2012}), simulated by reflecting
    $\rho_1$ from Fig.\,\ref{fig:main}A at the join. \cmnt{The model
      parameters match those in Fig.\,\ref{fig:main}C. Scale
      bars are 1\,mm.}}
  \label{fig:fgf8}
\end{center}

@@

The most involved part of this computation is to find the divergence of the
flux of axonal branching, $\mb{J}_i(\mb{x},t)$, the term in parentheses in
Eq.\,\ref{eq:da}:
%
\begin{equation}
  \label{eq:divJ}
  \nabla\vcdot\mb{J}_i(\mb{x},t) = \nabla\vcdot\left(D \nabla
  a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x}) \color{blue}+
  \frac{\epsilon a_i}{N-1} \nabla \hat{a}_i \color{black}\right),
\end{equation}
%
\cmnt{where we have defined $\hat{a}_i=\sum_{j\ne i}^{N}a_j$.} Note that the sum of the
guidance gradients is time-independent and define $\mb{g}_i(\mb{x}) \equiv
\sum_{j=1}^{M} \gamma_{i,j} \nabla\rho_j(\mb{x})$.  Because the divergence
operator is distributive, Eq.\,\ref{eq:divJ} can be expanded using vector
calculus identities (dropping references to $\mb{x}$ and $t$ for clarity):
%
\begin{equation}
\nabla\vcdot\mb{J}_i = \nabla\vcdot\big(D \nabla a_i\big) -
\nabla\vcdot\big(a_i \mb{g}_i\big) \color{blue} + \frac{\epsilon}{N-1}\nabla\vcdot\big({a}_i\nabla\hat{a}_i\big)\color{black}.
\end{equation}
%
Applying the vector calculus product rule identity yields
%
\begin{equation} \label{eq:divJExpanded}
%
\nabla\vcdot\mb{J}_i =
%
D \dvrg a_i % term1
-
a_i\nabla\vcdot\mb{g}_i % term2
-
\mb{g}_i\vcdot\nabla a_i % term3
\color{blue}
+ \frac{\epsilon a_i}{N-1} \dvrg \hat{a}_i % term1_1
+ \frac{\epsilon}{N-1} \nabla \hat{a}_i\vcdot\nabla{a}_i% term1_2
\color{black}
,
\end{equation}
%
which has \cmnt{five} elements to compute: i)~$D \dvrg a_i$ (the Laplacian of
$a_i$); ii)~a time-independent modulator of $a_i$ (because
$\nabla\vcdot\mb{g}_i$ is a time-independent static field); iii)~the scalar
product of the static vector field $\mb{g}_i$ and the gradient of $a_i$;
\cmnt{iv)~the Laplacian of $\hat{a}_i$; and v)~a term involving the gradients
  of $a_i$ and $\hat{a}_i$}. Each of the divergences can be simplified by
means of Gauss's Theorem following \cite{lee_hexagonal_2014}.

(i) The computation of the mean value of the Laplacian across one hexagon of
area $\Omega = \frac{\sqrt{3}}{2}d^2$, located at position $\mb{p}_0$, with
neighbours at positions $\mb{p}_1$--$\mb{p}_6$ is
%
\begin{equation} \label{eq:lapl}
\begin{split}
\langle D \dvrg a_i(\mb{p}_0,t) \rangle  = \frac{D}{\Omega} \oiint_\Omega \dvrg a_i(\mb{x}, t) \dif\Omega & = \frac{D}{\Omega} \oint \frac{\partial a_i}{\partial \hat{\mb{n}}} \dif\gamma \\
& \approx \frac{D}{\Omega} \sum_{j=1}^6 \frac{\partial a_i(\mb{p}_j)}{\partial \hat{\mb{n}}} \bigg\rvert_{\mathrm{mid}} v \\
& = \frac{2D}{\sqrt{3} d^2} \sum_{j=1}^6 \frac{a_i(\mb{p}_j) - a_i(\mb{p}_0)}{d} \frac{d}{\sqrt{3}} \\
& = \frac{2D}{3 d^2} \sum_{j=1}^6 \big(a_i(\mb{p}_j) - a_i(\mb{p}_0)\big),
\end{split}
\end{equation}
%
where $v = d/\sqrt{3}$ is the length of each edge of the hexagon and $\dif\gamma$
is an infinitesimally small distance along its perimeter.

\noindent
ii) The computation of the second term in Eq.\,\ref{eq:divJExpanded},
$\langle a_i(\mb{p}_0,t)\nabla\vcdot\mb{g}_i(\mb{p}_0)\rangle$, can be written out similarly:

@@

\noindent
iii) The \cmnt{middle} term in Eq.\,\ref{eq:divJExpanded} is the scalar product of two
vector fields which is straightforward to compute from their Cartesian
components.

\noindent
\cmnt{iv) The same method used to compute $\dvrg a_i$ in term (i) is used to compute
$\dvrg \hat{a}_i$.}

\noindent
\cmnt{v) The final term is the scalar product of the two vector fields $\nabla
  a_i$ and $\nabla \hat{a}_i$.}

By separating the computation of Eq.\,\ref{eq:divJ} into parts (i)--(v), the
no-flux boundary condition,
%
\begin{equation}
  \label{eq:noflux}
  \mb{J}_i(\mb{x},t)\big\rvert_{\mathrm{boundary}} = 0,
\end{equation}
%
can be fulfilled. On
the boundary, the contribution to $\mb{J}$ resulting from the first term of
Eq.\,\ref{eq:divJExpanded} can be fixed to 0 by the `ghost cell method' in
which, during the evaluation of (i), a hex outside the boundary containing the
same value as the hex inside the boundary is imagined to exist such that the
flux of $\mb{J}$ across the boundary is 0. Then, $\mb{g}_i(\mb{x})$ can be
tailored so that it, and its normal derivative approach 0 at the boundary,
ensuring that the second and third terms of Eq.\,\ref{eq:divJExpanded} also
contribute nothing to $\mb{J}$. This is achieved by \cmnt{multiplying}
$\mb{g}_i(\mb{x})$ \cmnt{by} a sharp logistic function of the distance, $d_b$,
from $\mb{x}$ to the boundary, \cmnt{of the form $1/[1 +
  \exp(100(d_f-d_b))]$, where $d_f=0.1$\,mm is the boundary fall-off distance,
  chosen to be approximately $3d$}.

All code required to reproduce these results is available at
\url{https://github.com/ABRG-Models/BarrelEmerge/tree/eLife}. The computations
described in (i)--(v) may be found in the class method
\code{RD\_James\cmnt{\_comp2}::compute\_divJ()} which calculates \code{term1},
\code{term2}, \code{term3}, \code{term1\_1} and \code{term1\_2}, respectively.

@@

Movie corresponding to Fig.\,\ref{fig:main}C in the main paper. Simulation parameters were
$N=41$, $\alpha=3.6$, $\beta=16.67$, $k=3$, $D=0.5$, $\gamma\in\pm 2$,
$\epsilon=1.2$ and $\delta{t}=0.0001$. Colours indicate the thalamic
projection for which the connection density is maximal, black lines delineate
boundaries, and overlaid contours show $c>0.5$. The final frame in the movie
is step $30000$ of the simulation.
